---
title: "Supplementary Figure 4"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# FigS4A+D

```{r}
library(xlsx)
library(reshape2)

results = read.delim("LMM_DFxR.txt")
df <- subset(results, results$Subset == "CD11c")

##  feature 1: LOGF2
    geneList <- df$Estimate

##  feature 2: named vector
    names(geneList) <- df$Gene

##  feature 3: decreasing order
    geneList <- sort(geneList, decreasing = TRUE)
    head(geneList)
    class(geneList) #should be numeric

##  Create a LogF2 cut-off e.g. > 1
    gene <- geneList[abs(geneList) > 0.5] #Change log2fold cutoff here
    head(gene)

####### Creating a custom pathway #############################################################
##  Make a tab-delimited file for your custom pathways
##  [First column: name of pathway][Following columns: genes belonging to pathway]
## !!! Make sure this is in the right species format !!!

    path <- read.xlsx("DC_combined.xlsx", sheetIndex = 1)
    
    path.melt = melt(t(path))
    
    path.melt = data.frame(signature = path$Signature
                           gene = path$Genes)

####### Running the GSEA #############################################################

    em <- GSEA(geneList = geneList,
                    exponent = 1, #standard
                    minGSSize = 10, #Minimum amount of genes in one geneset
                    maxGSSize = 1000, #Maximum amount of genes in one geneset
                    pAdjustMethod = "BH", 
                    TERM2GENE = path.melt, #genesets
                    pvalueCutoff = 1, #put 1 to see all pathways
                    verbose = TRUE)
 
##  Save your GSEA output
    setwd("YOUR_PATH") #where you will write your file

    head(as.data.frame(cbind(em@result$ID, em@result$pvalue, em@result$p.adjust, em@result$NES)))
    xy <- as.data.frame(cbind(em@result$ID, em@result$pvalue, em@result$p.adjust, em@result$NES))
    colnames(xy) <- c("Pathway", "P-value", "Adj. P-value", "NES")
    write.table(xy, "FILENAME.txt", sep = "\t", row.names = F, col.names = T)
``` 

library(xlsx)
library(reshape2)

results = read.delim("LMM_DFxR.txt")
df <- subset(results, results$Subset == "CD68")

##  feature 1: LOGF2
    geneList <- df$Estimate

##  feature 2: named vector
    names(geneList) <- df$Gene

##  feature 3: decreasing order
    geneList <- sort(geneList, decreasing = TRUE)
    head(geneList)
    class(geneList) #should be numeric

##  Create a LogF2 cut-off e.g. > 1
    gene <- geneList[abs(geneList) > 0.5] #Change log2fold cutoff here
    head(gene)

####### Creating a custom pathway #############################################################
##  Make a tab-delimited file for your custom pathways
##  [First column: name of pathway][Following columns: genes belonging to pathway]
## !!! Make sure this is in the right species format !!!

    path <- read.xlsx("Nat Comm 2024_TAM atlas.xlsx", sheetIndex = 1)
    
    path.melt = melt(t(path))
    
    path.melt = data.frame(signature = path$Signature
                           gene = path$Genes)

####### Running the GSEA #############################################################

    em <- GSEA(geneList = geneList,
                    exponent = 1, #standard
                    minGSSize = 10, #Minimum amount of genes in one geneset
                    maxGSSize = 1000, #Maximum amount of genes in one geneset
                    pAdjustMethod = "BH", 
                    TERM2GENE = path.melt, #genesets
                    pvalueCutoff = 1, #put 1 to see all pathways
                    verbose = TRUE)
 
##  Save your GSEA output
    setwd("YOUR_PATH") #where you will write your file

    head(as.data.frame(cbind(em@result$ID, em@result$pvalue, em@result$p.adjust, em@result$NES)))
    xy <- as.data.frame(cbind(em@result$ID, em@result$pvalue, em@result$p.adjust, em@result$NES))
    colnames(xy) <- c("Pathway", "P-value", "Adj. P-value", "NES")
    write.table(xy, "FILENAME.txt", sep = "\t", row.names = F, col.names = T)
```

The following document was created in Microsoft Excel, combining the GSEA output of all three celltypes comparing DF vs R.

```{r}
library(xlsx)
library(ggplot2)
library(ggpubr)

df <- read.xlsx("NESscores_signatures.xlsx", sheetIndex = 1)

df$LogP = -log(df$adj.P)
df$up = ifelse(df$NES < 0, "R", "DF")
df$NES = -1*df$NES
df$ID = paste0(df$NES,"_", df$Cluster.ID)

p1 <- ggballoonplot(
  subset(df, Sig.Select == "DC_combined"),
  x = "NES",
  y = "ID", facet.by = "Celltype",
  size = "LogP", fill = "NES", size.range = c(1,6)) +
  scale_fill_fermenter(n.breaks = 9, palette = "RdYlBu") +  # Set min and max for the fill color
  coord_fixed(ratio = 1)  # Adjust the ratio as needed

p2 <- ggballoonplot(
  subset(df, Sig.Select == "Nat Comm 2024_TAM atlas"),
  x = "NES",
  y = "ID", facet.by = "Celltype",
  size = "LogP", fill = "NES", size.range = c(1,6)) +
  scale_fill_fermenter(n.breaks = 9, palette = "RdYlBu") +  # Set min and max for the fill color
  coord_fixed(ratio = 1)  # Adjust the ratio as needed

p3 <- ggballoonplot(
  subset(df, Sig.Select == "Pan-cancer-TIL-atlas-Science_CD8-signatures"),
  x = "NES",
  y = "ID", facet.by = "Celltype",
  size = "LogP", fill = "NES", size.range = c(1,6), ) +
  scale_fill_fermenter(n.breaks = 9, palette = "RdYlBu") +  # Set min and max for the fill color
  coord_fixed(ratio = 1)  # Adjust the ratio as needed

print(p1)
print(p2)
print(p3)
```

# FigS4B

```{r}
library(xlsx)
library(pheatmap)
library(dplyr)
library(stringr)

averageCounts = read.delim("counts_Q3norm_uniqueSamples.txt", row.names = 1)

sub.count <- averageCounts[, grepl("CD11c", names(averageCounts))]

# Function to calculate z-score for a row
row_zscore <- function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}

# Apply the function to each row of the data frame
df_zscores <- t(apply(sub.count, 1, row_zscore))
df_zscores <- as.data.frame(df_zscores)

df <- data.frame(row.names = rownames(df_zscores),
                 DF = rowMeans(df_zscores[, grep("_DF", names(df_zscores))], na.rm = TRUE),
                 R = rowMeans(df_zscores[, grep("_R", names(df_zscores))], na.rm = TRUE))

sampleInfo = data.frame(SampleID = colnames(df))

sampleInfo <- sampleInfo %>%
  mutate(Celltype = case_when(
    grepl("CD68", SampleID) ~ "CD68",
    grepl("CD11c", SampleID) ~ "CD11c",
    grepl("CD8", SampleID) ~ "CD8",
    TRUE ~ NA_character_  # or you can leave it empty if you don't want NA values
  ))

sampleInfo <- sampleInfo %>%
  mutate(Group = case_when(
    grepl("_DF", SampleID) ~ "DF",
    grepl("_R", SampleID) ~ "R",
    TRUE ~ NA_character_  # or you can leave it empty if you don't want NA values
  ))

sampleInfo <- sampleInfo %>%
  mutate(PatientID = str_extract(SampleID, "P\\d{3}"))

saminf <- sampleInfo[, -1]
rownames(saminf) <- sampleInfo$SampleID

rownames(saminf) == colnames(df)

library(reshape2)
genes <- read.xlsx("FigS4_heatmap_gene_selection", sheetIndex = 1)
genes.melt <- melt(as.matrix(genes))
genes.melt <- na.omit(genes.melt)
colnames(genes.melt) <- c("x", "list", "gene")

library(pheatmap)
#Column annotation
my_sample_col <- data.frame(Group = saminf$Group,
                            Celltype = saminf$Celltype)

for (i in unique(genes.melt$list)[11:21]){
x <- subset(genes.melt, list == i)$gene
hm <- df[rownames(df) %in% x,]
p <- pheatmap(mat = hm, 
         cluster_rows = T, 
         cluster_cols = F, 
         #scale = "row", 
         fontsize = 7, 
         #annotation_col = my_sample_col,
         color = hcl.colors(100, "Mint", rev = T),
         #color = hcl.colors(100, "RdBu", rev = T),
         cutree_cols = NA, 
         cutree_rows = NA, angle_col = 45, fontsize_col = 9, main = paste0("DC - ", i), 
         cellwidth = 20,    # Set desired cell width
         cellheight = 20)
}
```
# FigS4C

Pathway analysis with [Metascape](https://metascape.org/gp/index.html)

# FigS4E

```{r}
library(xlsx)
library(pheatmap)
library(dplyr)
library(stringr)

averageCounts = read.delim("counts_Q3norm_uniqueSamples.txt", row.names = 1)

sub.count <- averageCounts[, grepl("CD68", names(averageCounts))]

# Function to calculate z-score for a row
row_zscore <- function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}

# Apply the function to each row of the data frame
df_zscores <- t(apply(sub.count, 1, row_zscore))
df_zscores <- as.data.frame(df_zscores)

df <- data.frame(row.names = rownames(df_zscores),
                 DF = rowMeans(df_zscores[, grep("_DF", names(df_zscores))], na.rm = TRUE),
                 R = rowMeans(df_zscores[, grep("_R", names(df_zscores))], na.rm = TRUE))

sampleInfo = data.frame(SampleID = colnames(df))

sampleInfo <- sampleInfo %>%
  mutate(Celltype = case_when(
    grepl("CD68", SampleID) ~ "CD68",
    grepl("CD11c", SampleID) ~ "CD11c",
    grepl("CD8", SampleID) ~ "CD8",
    TRUE ~ NA_character_  # or you can leave it empty if you don't want NA values
  ))

sampleInfo <- sampleInfo %>%
  mutate(Group = case_when(
    grepl("_DF", SampleID) ~ "DF",
    grepl("_R", SampleID) ~ "R",
    TRUE ~ NA_character_  # or you can leave it empty if you don't want NA values
  ))

sampleInfo <- sampleInfo %>%
  mutate(PatientID = str_extract(SampleID, "P\\d{3}"))

saminf <- sampleInfo[, -1]
rownames(saminf) <- sampleInfo$SampleID

rownames(saminf) == colnames(df)

library(reshape2)
genes <- read.xlsx("FigS4_heatmap_gene_selection", sheetIndex = 1)
genes.melt <- melt(as.matrix(genes))
genes.melt <- na.omit(genes.melt)
colnames(genes.melt) <- c("x", "list", "gene")

library(pheatmap)

#Column annotation
my_sample_col <- data.frame(Group = saminf$Group,
                            Celltype = saminf$Celltype)

for (i in unique(genes.melt$list)[1:10]){
x <- subset(genes.melt, list == i)$gene
hm <- df[rownames(df) %in% x,]
p <- pheatmap(mat = hm, 
         cluster_rows = T, 
         cluster_cols = F, 
         #scale = "row", 
         fontsize = 7, 
         #annotation_col = my_sample_col,
         color = hcl.colors(100, "Mint", rev = T),
         #color = hcl.colors(100, "RdBu", rev = T),
         cutree_cols = NA, 
         cutree_rows = NA, angle_col = 45, fontsize_col = 9, main = paste0("Mac - ", i), 
         cellwidth = 20,    # Set desired cell width
         cellheight = 20)
}
```

# FigS4C

Pathway analysis with [Metascape](https://metascape.org/gp/index.html)